---
title: "EACE Extremity Trauma Outcomes — Exploratory Data Analysis"
subtitle: "Fellowship ID: EACE-2026-0006"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  # pdf_document: default   # uncomment for PDF output
params:
  data_path: "../outputs/features.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 8)
library(tidyverse)
library(table1)

df <- read_csv(params$data_path, show_col_types = FALSE)

# Harmonize expected column names
if (!"Cohort" %in% names(df) && "Cohort_Type" %in% names(df)) {
  df <- df %>% rename(Cohort = Cohort_Type)
}
df$Cohort <- factor(df$Cohort, levels = c("Salvage", "Amputation"))
```

## 1. Cohort Overview

```{r cohort-counts}
df %>%
  distinct(Patient_ID, Cohort) %>%
  count(Cohort) %>%
  knitr::kable(caption = "Patients per cohort")
```

## 2. Table 1 — Baseline Demographics

Baseline (first-record) characteristics stratified by cohort assignment.
Continuous variables reported as Mean (SD); categorical as N (%).

```{r table1, results='asis'}
baseline <- df %>%
  group_by(Patient_ID) %>%
  slice_min(Days_Since_Injury, n = 1, with_ties = FALSE) %>%
  ungroup()

label(baseline$Age) <- "Age (years)"
label(baseline$ISS_Score) <- "Injury Severity Score"
label(baseline$BMI) <- "BMI"
label(baseline$Mechanism_of_Injury) <- "Mechanism of Injury"

table1(~ Age + ISS_Score + BMI + Mechanism_of_Injury + Pre_Existing_Diabetes | Cohort,
       data = baseline, overall = "Total")
```

## 3. Functional Recovery Trajectories (SMFA Mobility)

```{r smfa-trajectory}
df %>%
  group_by(Cohort, Days_Since_Injury) %>%
  summarise(mean_smfa = mean(SMFA_Mobility, na.rm = TRUE),
            se = sd(SMFA_Mobility, na.rm = TRUE) / sqrt(n()), .groups = "drop") %>%
  ggplot(aes(x = Days_Since_Injury, y = mean_smfa, color = Cohort, group = Cohort)) +
  geom_line(linewidth = 1) +
  geom_ribbon(aes(ymin = mean_smfa - se, ymax = mean_smfa + se, fill = Cohort),
              alpha = 0.15, color = NA) +
  labs(title = "SMFA Mobility Score Over Time by Cohort",
       x = "Days Since Injury", y = "SMFA Mobility (0-100, higher = better)") +
  theme_minimal(base_size = 13)
```

## 4. Kaplan-Meier: Time to Return-to-Duty (or censoring)

```{r km-curve}
library(survival)
library(survminer)

rtd_df <- df %>%
  distinct(Patient_ID, Cohort, Return_to_Duty, Days_Since_Injury) %>%
  group_by(Patient_ID) %>%
  summarise(Cohort = first(Cohort),
            event = as.integer(any(Return_to_Duty == TRUE)),
            time = max(Days_Since_Injury), .groups = "drop")

fit <- survfit(Surv(time, event) ~ Cohort, data = rtd_df)
ggsurvplot(fit, data = rtd_df, risk.table = TRUE, conf.int = TRUE,
           legend.title = "Cohort", xlab = "Days Since Injury",
           ylab = "Proportion Not Yet Returned to Duty",
           title = "Time to Return-to-Duty by Cohort")
```

## 5. Correlation Heatmap: Psychological x Functional Outcomes

```{r corr-heatmap}
corr_vars <- df %>%
  select(SMFA_Mobility, SMFA_Emotional, PTSD_Checklist, Gait_Symmetry_Index,
         Daily_Step_Count) %>%
  drop_na()

corr_mat <- cor(corr_vars, use = "pairwise.complete.obs")

corr_long <- as.data.frame(as.table(corr_mat))

ggplot(corr_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), size = 3.5) +
  scale_fill_gradient2(low = "#2c7fb8", mid = "white", high = "#d7301f",
                        midpoint = 0, limits = c(-1, 1), name = "r") +
  labs(title = "Correlation: Psychological & Functional Outcomes", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 6. Notes

- Report auto-generated by `eda_report.Rmd`. Re-knit after any Module 1/2 pipeline
  change: `rmarkdown::render("eda_report.Rmd", params = list(data_path = "..."))`.
- Table 1 and KM curves use first-available and last-available follow-up records,
  respectively; adjust `slice_min`/`summarise` logic if your registry extract
  has a different visit cadence.
